The Folly of the Crowd: \\ Market Inefficiencies in Peer to Peer Lending

### 2015-03-28 draft.Rnw
added ds.tab. Softened EMH language

\documentclass[12 pt]{article}
\usepackage{graphicx, verbatim}
\usepackage{float}
\usepackage{subfig}
\usepackage{url}
\setlength{\textwidth}{7in} 
\setlength{\textheight}{9in}
\setlength{\oddsidemargin}{0in} 
\setlength{\evensidemargin}{0in}
\setlength{\topmargin}{-1.5cm}
\overfullrule=2cm
\usepackage[linktoc=all]{hyperref}
\usepackage{titlesec}
\hypersetup{colorlinks, citecolor=black, filecolor=black, linkcolor=black,urlcolor=black}
\usepackage[citestyle=verbose,bibstyle=numeric,sorting=ynt,backend=biber]{biblatex}
\addbibresource{lit.bib}
\begin{document}
\setcounter{tocdepth}{2}
\linespread{1.3}
\tableofcontents

<<init,echo=FALSE,warning=FALSE,results='hide',message=FALSE,cache=FALSE,eval=TRUE>>=
#DONT CACHE SAYS YIHUI
require(dplyr)              ### Used Extensively
require(ggplot2)            ### Used Extensively
require(gridExtra)          ### For facet plots
require(knitr)              ### For chunk control
require(leaps)              ### For feature selection
require(lubridate)          ### For date grouping, conversions
require(reshape2)           ### For melt()
require(scales)             ### For plotting
require(texreg)             ### Regression Output using texreg()
require(tidyr)              ### For spread()
require(xtable)             ### Tabular Output using xtable()
options(scipen=999,digits=4,formatR.arrow=TRUE,width=90)
coruse = 'pairwise.complete.obs'
opts_chunk$set(list(fig.path='figure/minimal-', fig.align='center',echo=FALSE, fig.show='hold',tidy=TRUE,fig.height=5,fig.pos='H',warning=FALSE, message=FALSE,results='asis'))
@

@@

\title{The Folly of the Crowd: \\ Market Inefficiencies in Peer to Peer Lending}
\author{Samuel Shleifer\\ Yale University}
\date{\today}
\maketitle 
\begin{abstract}
This paper documents violations of market efficiency in Peer to Peer Lending, using data from Prosper.com, America's first Peer-to-Peer Lending Marketplace. Peer to Peer lending is the practice of individual and institutional investors lending money to borrowers over the internet, without the assistance of a traditional financial intermediary. Theories of Market Efficiency predict that investors price assets to earn an expected return commensurate with risk taken, and that asset ``prices should fully reflect publicly available information." Contrary to these, and other, efficient market hypothesis predictions: 
\begin{enumerate}
\item Investors in risky Prosper loans earn negative average returns during the first few years of its history.
\item Investors earn lower returns on riskier loans during the first few years of Prosper's history.
\item Strategies using publicly available information can earn superior risk-adjusted returns.
\item Poor loan performance during the first few years of Prosper's history is a consequence of poor, and not just an unfortunate time period. Had investors better used the information at their disposal, they would have earned better returns.
\end{enumerate}
These violations of market efficiency are discussed in the midst of two market design changes by Prosper that guided and then forced investors towards setting more efficient interest rates. Such guidance and eventual price control, as opposed to unconstrained auction rate setting, improved loan performance.
\end{abstract}
\newpage
\section{Introduction}
Historically, there are two major perspectives of efficient market prices. Fama ((1970)'s semi-strong efficient markets hypothesis ("EMH") predicts that a strategy using publicly available information should not be able to earn excess risk-adjusted returns. Risk models, such as CAPM, imply that risky assets should have positive expected returns and that riskier assets should have higher returns than less risky assets.  The information aggregation perspective associated with (Hayek, 1945) and (Stiglitz and Grossman, 1980), further predicts that many individuals, each using their own information, collectively arrive at a more efficient price than a central planner who sets prices, since the central planner cannot access all the information available to the crowd.\footcite{stiglitz}\footcite{hayek}

In the P2P Lending context, the market's price is the interest rate associated with a loan. This paper studies the question of whether interest rates on Prosper.com are set efficiently. (Fama, 1970)'s semi-strong EMH predicts that if interest rates were set efficiently, then...
\begin{enumerate}
\item Loans would have positive expected returns. Investors would be unwilling to lend to Borrowers if they expected to lose money. \footnote{Assuming a Risk-Free Rate of 0}
\item Riskier loans would have higher expected returns. \footcite{sharpe}
\item It is not possible to do better than the average investor using publicly available information. \footcite{fama}
\end{enumerate}
I test hypothesis 1 by measuring the historical returns of loans up to a certain date, and then asking whether loans with negative expected returns get funded at that date.
I test hypothesis 2 by comparing the Returns of high risk loans to low risk loans.
I test hypothesis 3 by using publicly available information to select loans with a higher than average expected return.

@@

The early auction regime is closest to Hayek's vision of the free market, since interest rates are determined freely by investors through auctions.  In this case, the EMH predictions of no negative average returns, a positive relationship between risk and return, and no improvement in performance from using public information are most applicable.  Yet, as I show, it is precisely in this regime that the three predictions of the EMH are most dramatically violated. In the two later regimes, when Prosper first provides much more guidance to investors, and then actually sets interest rates, investor returns improve dramatically. The evidence across regimes is broadly inconsistent with the view that local knowledge and the wisdom of the crowd improves price discovery. To the contrary, it appears that a "centralized' Prosper mechanism of either guiding or setting interest rates actually leads to fewer predictable mispricings: as Prosper becomes more involved in the rate-setting process, the excess returns available using publicly available information approach 0, while aggregate returns exhibit higher mean and lower variance.

@@

@@

Under all three regimes, lenders could choose to whom to lend and how much to lend, allowing them to diversify across many loans, thereby decreasing the variance of their returns. In exchange for verifying borrowers' identities and managing servicing, Prosper charges a 0.5-5 percentage point origination fee on each loan depending on the borrower's risk profile and loan duration. This fee is usually 1\%. Fees are added to the borrower's monthly payment.\footnote{There are no prepayment penalties.} This rate is lower than that charged by most traditional financial institutions, which have the higher costs of physical branches, and pay for human analysis of credit history.\footcite{hanson}
The P2PL model relies on lenders being able to infer the riskiness of borrowers: since Prosper gets fees from monthly payments, they too are incentivized to maximize investor returns, both for short term profit, and long term reputation.\newline

@@

\subsection{Auction Design}
From April 19, 2006 until December 20, 2010. Prosper operated a \emph{variable rate} model, where lenders and borrowers determined interest rates using a Dutch auction-like system, where each investor bid consisted of a (dollar amount, interest rate) pair. In these auctions, borrowers would declare a maximum reservation interest rate, where bidding would start. Over the course of a 7-10 day bidding period, that rate would be bid down by investors, with the loan's interest rate closing at the lowest rate $R_{clearing}$ such that enough investors were willing to fully fund the loan at or below the $R_{clearing}$.All investors who bid at or below the clearing rate would receive the clearing rate. If a borrower could not reach 100\% funding at his reservation rate, the loan would not be funded.\newline
As an instructional example, suppose Bill the borrower needed to borrow \$100, and set his maximum rate at 20\%.
Let Chris, David and Ellen be investors, with Chris bidding (\$50, 15\%), David bidding (\$50, 18\%) and Ellen bidding (\$30,19\%). In this scenario, Chris and David would win the auction, and each lend lend Bill \$50, and receive an 18\% interest rate. Bill would pay 18\% interest before fees.\newline

\subsection{Early Auction Regime}
In the early auction regime, loans originated between April 19, 2006 and October 15, 2008, auction participants had access to considerable information from Prosper, including over 60  Borrower characteristics, such as residence, employment status and a FICO score bucket, but few intelligent risk estimates. Additionally, Prosper did not regulate who could apply for a loan on the website. Besides basic fraud prevention, any applicant who received 100\% funding from investors was eligible, for Prosper's first year of operation. On March 1, 2007, amidst poor returns, Prosper announced that listings with credit scores below 520 would no longer be offered to investors. This intervention is not discussed for lack of space, but shows that after a year, more or less, Prosper felt the need to nudge a fully decentralized investment process in the right direction.\newline
In an  October 15, 2008 email, Prosper announced that it would temporarily stop accepting new loan applications to register with the SEC, which subsequently released a cease and desist order characterizing Prosper as a ``seller of securities," rather than a marketplace merely matching lenders and borrowers, as the site had argued.
\footnote{\url{techcrunch.com/2008/11/26/sec-outlines-its-reasoning-for-shutting-down-p2p-lender-prosper/}}\footnote{SEC LETTER}\newline
\subsection{Late Auction Regime}
After a 9 month quiet period, Prosper re-opened, using the same Auction model, with three significant market interventions. First, under-writing standards again became more stringent: new borrowers were required to have a FICO score above 640, and repeat borrowers were required to have a score above 600.

@@

\subsection{Soft Information and Credit Screening}
The novel idea of P2P is a large group of investors may be better equipped to assess risk than traditional financial institutions. ``Traditionally, the credit score provided by rating agencies has been the main tool banks use to screen smaller borrowers."\footcite{iyer} As such, the initial strand of literature focuses largely on the novel sources of information utilized by P2P investors.
(Gao and Lin, 2012) analyze the text written by the borrower in the ``description field" of the application, and find that descriptions that are easy to read have 2.3\% lower default rates than average, while complex narratives are associated with a 3.6\% higher default rate, but do not analyze return on investment.\footcite{gao} Pope and Sydnor (2008) find that investors can reduce their defaults by statistically discriminating against racial minorities, but do not analyze return on investment. (Iyer et. al,2009) find that within the credit score buckets provided by Prosper, investors use soft information to further sort borrowers within buckets, reducing defaults.\footcite{iyer}
These papers echo Hayek (1945)'s basic argument for market efficiency: that many individuals, each using their own information, collectively arrive at one efficient price.  The right price ``might have been arrived at by one single mind possessing all the information which is in fact dispersed among all the people involved in the process." \footcite{hayek}

This essay contributes to this literature along two dimensions. First, I use ROI, rather than a simple default rate, to measure the improvement of investment strategies over a benchmark. Second, and more importantly, my paper pushes back against the idea that the soft information used by borrowers is more useful for credit screening; I find that borrowers do not do a very good job aggregating the various hard credit information at their disposal, and synthesizing this into an interest rate that reflects the riskiness of the borrower. When Prosper, as a central planner, assists them, they set much more efficient prices.
%Liquidity,cost-induced default and other unused citations can be found in hypo.cut.tex

\section{Data}
My sample consists of 39,555 three year, unsecured (no collateral) Loans, worth \$231.7 Million,  originated from 2006-2011, downloaded off Prosper.com at various points by various people. Borrowers apply for loans by creating a public listing on the Prosper.com website, and can request anywhere from \$1,000 to \$35,000.\footnote{The minimum increased to \$2,000 effective Nov. 1, 2011}. Since our analysis requires measurement of loan performance, which can only take place 3 years after origination(given 3 year maturity), we must end the sample in 2011, as we do not know what will end up happening to more recently originated loans. As shown in Table \ref{sample}, this leaves us with many more early auction regime loans than the post shutdown regimes.\newline
<<sampTab, cache=FALSE>>=
xtable(count(nd,regime) %>% transform(freq. =  n/nrow(nd)), 
       digits=4, caption='Observations per regime. Each observation is one loan.', label='sample')
@

To help investors better assess the creditworthiness of borrowers, for every loan originated in the site's history, Prosper posts data regarding loan performance, such as payments made, interest rate, and principal lost. Borrower characteristics, including credit variables, which are usually obtained directly from a consumer credit agency like FICO, and verification of income and employment, are also posted on the website in the borrower's \emph{listing} for prospective investors to consider as they decide who to lend to, and available for download.
\subsection{Variable Definitions for the Prosper Data}
<<sumTable,cache=FALSE>>=
nd$default.rate =  nd$defaulted
k4 = c('roi','default.rate', 'LenderYield','BorrowerRate','regime','LoanOriginalAmount','Investors','CreditScoreEst','DebtToIncomeRatio', 'InquiriesLast6Months')
ag4 <- aggregate(.~regime, nd[k4], function(x) c(mean = mean(x), sd = sd(x)))
d4 = t(do.call(data.frame, c(ag4, check.names = FALSE)))
d5 = data.frame(d4)
names(d5) = d4[1,]
d6 = d5[-1,]
xtable(d6,caption='Means and Standard Deviations for Relevant Variables by Regime.',digits=3, label='sumStats')
@
For each loan, Prosper offers 63 interesting variables, I now define an important subset.
\begin{itemize}
\item BorrowerRate: Annual interest rate paid by the borrower
\item TotalCreditLinespast7years: Number of credit lines, including cards and loans, used by the borrower in the seven years before the prosper loan was originated.
\item Inquirieslast6Months: Number of times a lender inquired about the Borrower's credit report, in the 6 months before the loan was originated. Traditional lenders make inquiries when a borrower applies for a loan, so this serves as a proxy for the number of times someone has borrowed in the past 6 months.
\item CurrentDelinquencies: Number of loans on which the borrower is delinquent at the origination date.
\item DelinquenciesLast7Years: The number of loans which the borrower has not repaid fully over the 7 years before the origination date.
\item MonthlyLoanPayment: The borrower's scheduled monthly payment on the Prosper loan.
\item CreditScoreRangeLower: The lower bound of the credit score range (which bounds his FICO score) , that bounds a borrower's credit grade, as provided by FICO.
\item CreditScoreEst: The average of the lower and upper bounds of the credit score range. 
\item PublicRecordsLast10Years: Number of public records of borrowing activity in the past 10 years at the origination date. 
\item DebtToIncomeRatio: The debt to income ratio of the borrower at the origination date. This value is capped at 10.01 (any debt to income ratio larger than 1000\% will be coded as 1001\%).
\item TotalProsperPaymentsBilled: Number of on time payments the borrower made on Prosper loans at the time they created the listing. This value is 0 if the borrower had no prior loans.
\item LenderYield: The lender's Yield on the loan. Equal to the Interest rate minus servicing fee.
\item BorrowerRate: The borrower's interest rate for the loan.
\item Investors: The number of investors that funded the loan.
\item *.sc: The Variable * was scaled to have 0 mean and unit variance. (found only in regression results)
\end{itemize}

@@

I do not discount or annualize ROIs, because in the period of low interest rates, the difference between default and repayment is large enough that a few percentage points don't matter: loans that default average -46\% ROI, loans that repay average + 24\% ROI.\footnote{I will probably use discounting in the final draft, but it doesn't change results, and it feels like a low priority, at this stage.}
\footnote{Elaborate}

\subsection{Prosper's Charged-Off vs. Default Terminology}
After the shutdown, Prosper began using the term ``Charged-Off" to mean that a loan is delinquent past 120 days. We observe more defaults than charge-offs before the shutdown, but more charge-offs than defaults after, likely because of the shift in terminology. According to Prosper's website:
\begin{quote}
In general, a debt or account is considered “charged off” when it is unlikely that further payments will be received. Debts are usually charged off after they remain unpaid for a period of time (e.g., 90 to 180 days). Prosper uses the 120 days as the charge off threshold because loans that become over 120 days past due are eligible for sale to a debt buyer, and we have found that there is a steep drop-off in likelihood of further payments after 120 days of delinquency.
\footnote{\url{http://blog.prosper.com/2008/09/16/charge-offs-explained/}}
\end{quote}

\section{Performance Across Regimes}
This section begins with an examination of ROIs over time and and across borrower credit buckets. I then discuss regime differences in default rates and interest rates, the two features that largely determine returns. The discussion of interest rates includes measurement of investors' varying sensitivity to risk across regimes. The section closes by examing the impact of Prosper's interventions, which reduce the valid borrower pool to a less-risky subset and assign borrowers more informative credit grades, on investor returns.

@@

Finding (1) is inconsistent with the EMH prediction that risky assets like Prosper loans should offer investors a positive expected return.\footnote{Otherwise, investors would just hold cash.}
Finding (2) is inconsistent with the prediction that riskier loans should have higher expected returns to compensate investors for taking risks.
Finding (3) describes the sub-optimal investor decisions that led to (1) and (2), in response to arguments that investors chose an unlucky time period.
Finding (4) is inconsistent with Hayek's notion that a free market will arrive at accurate prices without the meddling of a central planner.

>>>>

\documentclass[12 pt]{article}
\usepackage{graphicx, verbatim}
\usepackage{float}
\usepackage{subfig}
\usepackage{url}
\setlength{\textwidth}{7in} 
\setlength{\textheight}{9in}
\setlength{\oddsidemargin}{0in} 
\setlength{\evensidemargin}{0in}
\setlength{\topmargin}{-1.5cm}
\overfullrule=2cm
\usepackage[linktoc=all]{hyperref}
\usepackage{titlesec}
\hypersetup{colorlinks, citecolor=black, filecolor=black, linkcolor=black,urlcolor=black}
\usepackage[citestyle=verbose,bibstyle=numeric,sorting=ynt,backend=biber]{biblatex}
\usepackage[toc,page]{appendix}
\addbibresource{lit.bib}
\begin{document}
\setcounter{tocdepth}{2}
\linespread{1.3}
\tableofcontents

<<init,echo=FALSE,warning=FALSE,results='hide',message=FALSE,cache=FALSE,eval=TRUE>>=
#DONT CACHE SAYS YIHUI
require(dplyr)              ### Used Extensively
require(ggplot2)            ### Used Extensively
require(gridExtra)          ### For facet plots
require(knitr)              ### For chunk control, compilation
require(leaps)              ### For feature selection
require(lubridate)          ### For date grouping, conversions
require(reshape2)           ### For melt() before ggplot
require(scales)             ### For plots
require(texreg)             ### Regression Output using texreg()
require(tidyr)              ### For spread()
require(xtable)             ### Tabular Output using xtable()
options(scipen=999,digits=4,formatR.arrow=TRUE,width=90)
coruse = 'pairwise.complete.obs'
opts_chunk$set(list(fig.path='figure/minimal-', fig.align='center',echo=FALSE, fig.show='hold',tidy=TRUE,fig.height=5,fig.pos='H',warning=FALSE, message=FALSE,results='asis'))
@

@@

\title{The Folly of the Crowd: \\ Market Inefficiencies in Peer to Peer Lending}
\author{Samuel Shleifer\footnote{R code available at \url{github.com/sshleifer/ProsperThesis}}\\ Yale University}
\date{\today}
\maketitle 
\begin{abstract}
This paper documents violations of market efficiency in Peer to Peer Lending, using data from Prosper.com, America's first Peer-to-Peer Lending Marketplace.
Peer to Peer lending is the practice of individual and institutional investors lending money to borrowers over the internet, without the assistance of a traditional financial intermediary.
Theories of Market Efficiency predict that asset ``prices should fully reflect publicly available information." 
Contrary to this prediction, strategies using publicly available information can
earn superior risk-adjusted returns, when compared with the other loans.

This violation of market efficiency, and poor investor returns in general, are
discussed in the midst of two market design changes by Prosper that guided, and
then forced investors towards setting more efficient interest rates. Such
guidance and eventual price control, as opposed to the unconstrained market
design, where interest rates were set by auction, is associated with improved
investor returns and more efficient markets. The aforementioned strategies can
only earn excess returns in the first two regimes, before Prosper began directly
setting interest rates.
\end{abstract}
\newpage

\section{Introduction}
Historically, there are two major perspectives of efficient market prices. Fama
((1970)'s semi-strong efficient markets hypothesis ("EMH") predicts that a
strategy using publicly available information should not be able to earn excess
risk-adjusted returns. Risk models, such as CAPM, imply that risky assets should
have positive expected returns and that riskier assets should have higher
returns than less risky assets. The information aggregation perspective
associated with (Hayek, 1945) and (Stiglitz and Grossman, 1980), further
predicts that many individuals, each using their own information, collectively
arrive at a more efficient price than a central agent who sets prices, since the
agent cannot access all the information available to the crowd.\footcite{stiglitz}\footcite{hayek}

In the P2P Lending context, the market's price is the interest rate associated with a loan. This paper studies the question of whether interest rates on Prosper.com are set efficiently. (Fama, 1970)'s semi-strong EMH predicts that if interest rates were set efficiently, then...
\begin{enumerate}
\item Loans would have positive expected returns. Investors would be unwilling to lend to Borrowers if they expected to lose money. \footnote{Assuming a Risk-Free Rate of 0}
\item Riskier loans would have higher expected returns. \footcite{sharpe}
\item It is not possible to do better than the average investor using publicly available information. \footcite{fama}
\end{enumerate}
I test hypothesis 1 by measuring the historical returns of loans up to a certain date, and then asking whether loans with negative expected returns get funded at that date.
I test hypothesis 2 by comparing the returns of high risk loans to low risk loans.
I test hypothesis 3 by using publicly available information to select loans with a higher than average expected return.

@@

The early auction regime is closest to Hayek's vision of the free market, since interest rates are determined freely by investors through auctions.
I find that it is precisely in this free-market regime that the three
predictions of the EMH are contradicted.
In the two later regimes, when Prosper first provides much more guidance to investors, and then actually sets interest rates, investor returns improve dramatically.
The evidence across regimes is broadly inconsistent with the view that local knowledge and the wisdom of the crowd improves price discovery.
To the contrary, it appears that a "centralized' Prosper mechanism of either guiding or setting interest rates actually leads to fewer predictable mispricings: as Prosper becomes more involved in the rate-setting process, the excess returns available using publicly available information approach 0, while aggregate returns exhibit higher mean and lower variance.

@@

This hypothetical method suffers from the problem that it is impossible to
perfectly simulate the effect of charging risky borrowers higher interest rates
during a turbulent macroeconomic period, as Section 6 attempts.
A perfect dataset would have Prosper randomly assign different rate-setting
mechanisms to borrowers, rather than assigning all borrowers in a given time
period the same rate-setting mechanism, and then switching it when things aren't
going well.

In this spirit, I do not claim that poor investor decision making is the only
factor causing poor returns in the Early auction period, or that the impact of
the policy changes can be accurately isolated by my methodologies.
The contradictions with efficiency hypotheses 1 and 2 can certainly be partially
explained by the same bad luck that explains poor returns on all risky assets
during economically turbulent times.
I merely argue that returns in the early auction period were likely made worse
by poor investor decisions, as evidenced by the predictable mispricings that can
be exploited, as they are in Section 5, the core of my market inefficiency
argument, which finds that over a somewhat longer period of time, simple loan selection strategies can use publicly available information to earn higher returns than the benchmark.\footnote{Benchmark calculated by assuming \$1 investment in every available loan.}

@@

Under all three regimes, lenders could choose to whom to lend and how much to
lend, allowing them to diversify across many loans, thereby decreasing the
variance of their returns. Before a potential borrower's listing is posted on
the website, Prosper authenticates their social security number, driver's
license and address, and pulls the borrower's credit history from Experian,
which includes credit score, total number of delinquencies, current
delinquencies, and many other historical credit variables.\footcite{freedman} 
In exchange, In exchange for verifying borrowers' identities and managing the
marketplace, Prosper charges a 0.5-3 percentage
point origination fee on each loan depending on the borrower's risk profile and
loan duration. 
This fee,usually 1\%, is added to the borrower's monthly payment.
Lenders are also charged a servicing fee of 0.5-1\%. There are no prepayment penalties.
This rate is lower than that charged by most traditional financial institutions, which have the higher costs of physical branches, and pay for human analysis of credit history.\footcite{hanson}
The P2PL model relies on lenders being able to infer the riskiness of borrowers: since Prosper gets fees from monthly payments, they too are incentivized to maximize investor returns, both for short term profit, and long term reputation.\newline

@@

\subsection{Auction Design}
From April 19, 2006 until December 20, 2010. Prosper operated a \emph{variable
rate} model, where lenders and borrowers determined interest rates using a Dutch
auction-like system, where each investor bid consisted of a (dollar amount,
interest rate) pair, such that the dollar amount is greater than \$50, the
minimum bid, and the rate is capped at the borrower's reservation
rate.\footcite{freedman}
In these auctions, borrowers would declare a maximum reservation interest
rate, capped at 36\%, where bidding would start. Over the course of a 7-10 day bidding period, that rate would be bid down by investors, with the loan's interest rate closing at the lowest rate $R_{clearing}$ such that enough investors were willing to fully fund the loan at or below the $R_{clearing}$. All investors who bid at or below the clearing rate would receive the clearing rate. If a borrower could not reach 100\% funding at his reservation rate, the loan would not be funded.\newline
As an instructional example, suppose Bill the borrower needed to borrow \$100, and set his maximum rate at 20\%.
Let Chris, David and Ellen be investors, with Chris bidding (\$50, 15\%), David bidding (\$50, 18\%) and Ellen bidding (\$30,19\%). In this scenario, Chris and David would win the auction, and each lend lend Bill \$50, and receive an 18\% interest rate. Bill would pay 18\% interest before fees.\newline

\subsection{Early Auction Regime}
In the early auction regime, loans originated between April 19, 2006 and October 15, 2008, auction participants had access to considerable information from Prosper, including over 60  Borrower characteristics, such as residence, employment status and a FICO score bucket, but few intelligent risk estimates. Additionally, Prosper did not regulate who could apply for a loan on the website. Besides basic fraud prevention, any applicant who received 100\% funding from investors was eligible, for Prosper's first year of operation. On March 1, 2007, amidst poor returns, Prosper announced that listings with credit scores below 520 would no longer be offered to investors. This intervention is not discussed for lack of space, but shows that after a year, more or less, Prosper felt the need to nudge a fully decentralized investment process in the right direction.\newline
In an  October 15, 2008 email, Prosper announced that it would temporarily stop accepting new loan applications to register with the SEC, which subsequently released a cease and desist order characterizing Prosper as a ``seller of securities," rather than a marketplace merely matching lenders and borrowers, as the site had argued.
\footnote{\url{techcrunch.com/2008/11/26/sec-outlines-its-reasoning-for-shutting-down-p2p-lender-prosper/}}\footnote{SEC LETTER}\newline
\subsection{Late Auction Regime}
After a 9 month quiet period, Prosper re-opened, using the same Auction model,
with three significant market interventions. First, under-writing standards
again became more stringent: new borrowers were required to have a credit score score above 640, and repeat borrowers were required to have a score above 600.

@@

\subsection{Soft Information and Credit Screening}
The novel idea of P2P is a large group of investors may be better equipped to assess risk than traditional financial institutions. ``Traditionally, the credit score provided by rating agencies has been the main tool banks use to screen smaller borrowers."\footcite{iyer} As such, the initial strand of literature focuses largely on the novel sources of information utilized by P2P investors.
(Gao and Lin, 2012) analyze the text written by the borrower in the ``description field" of the application, and find that descriptions that are easy to read have 2.3\% lower default rates than average, while complex narratives are associated with a 3.6\% higher default rate, but do not analyze return on investment.\footcite{gao} Pope and Sydnor (2008) find that investors can reduce their defaults by statistically discriminating against racial minorities, but do not analyze return on investment. (Iyer et. al,2009) find that within the credit score buckets provided by Prosper, investors use soft information to further sort borrowers within buckets, reducing defaults.\footcite{iyer}

These papers echo Hayek (1945)'s basic argument for market efficiency: that many individuals, each using their own information, collectively arrive at one efficient price.  The right price ``might have been arrived at by one single mind possessing all the information which is in fact dispersed among all the people involved in the process." \footcite{hayek}

Freedman and Jin (2008) are more focused on the failures of the auction regime,
Although the main focus of their paper is the role of soft information and
social networks on Prosper, they also study how investors learn to weight these
different variables, and confirm the finding of negative expected returns in the
early auction regime: ``lenders, especially those that joined
Prosper early, did not fully understand the market risk. We estimate, on
average, Lenders would have expected an internal rate of return of -.62\% to
1.38\% if they had correctly understood the risk distribution of Prosper
loans.''``Vigorous'' lender learning, they argue, explains the increase in interest
rates and borrower creditworthiness, and associated increase in lender
returns.\footcite{freedman}

This essay continues Freedman and Jin's critical look at investor decisions in the early
auction regime, but with the benefit of three more years of data, and two more
major Prosper policy changes: informative grading and centralized rate-setting.
Instead of a focus on soft information, I describe the hard information that
went largely ignored by investors until Prosper aggregated it into more
informative credit grades. Instead of investor learning as the catalyst that led
to a more efficient market, I frame Prosper's rule changes as catalysts that
force investors to make better decisions.

Rather differently than Hayek, I find that when investors are left on their own
to set prices, markets are not nearly as efficient as the prices set by ``a
single mind'', that aggregates the data. Furthermore, as interest rate setting
becomes more centralized, interest rates are more sensitive to the hard credit variables
that actually predict default.
t and other unused citations can be found in hypo.cut.tex

\section{Data}
My sample consists of 39,555 three year, unsecured (no collateral) Loans, worth
\$231.7 Million,  originated from 2006-2011. The data was downloaded off Prosper.com.
Borrowers apply for loans by creating a public listing on the Prosper.com website, and can request anywhere from \$1,000 to \$35,000.\footnote{The minimum increased to \$2,000 effective Nov. 1, 2011}.
Since our analysis requires measurement of loan performance, which can only take
place three  years after origination (given 3 year maturity on all loans), the
sample is ended in 2011, as we do not know what will end up happening to more recently originated loans. 
As shown in Table \ref{sample}, this leaves us with many more early auction regime loans than the post shutdown regimes.\newline
<<sampTab, cache=FALSE>>=
xtable(count(nd,regime) %>% transform(freq. =  n/nrow(nd)), 
       digits=4, caption='Observations per regime. Each observation is one loan.', label='sample')
@

To help investors better assess the creditworthiness of borrowers, for every
loan originated in the site's history, Prosper posts historical data regarding loan
performance, such as payments made, interest rate, and principal lost. Borrower
characteristics, including credit variables, which are usually obtained directly
from Experian and verification of income and employment, are available to
investors for each listing. Moreover, historical market data is available for
download, through a query tool, and through third party websites. 
Interviews conducted at the 2008 Prosper Days Conference suggest that there was
``enormous heterogeneity in lender awareness of the data, ability to process the
data, and intent to track the data over time.''\footcite{freedman}

\subsection{Variable Definitions for the Prosper Data}
<<sumTable,cache=FALSE>>=
nd$default.rate =  nd$defaulted
k4 = c('roi','default.rate', 'LenderYield','BorrowerRate','regime','LoanOriginalAmount','Investors','CreditScoreEst','DebtToIncomeRatio', 'InquiriesLast6Months')
ag4 <- aggregate(.~regime, nd[k4], function(x) c(mean = mean(x), sd = sd(x)))
d4 = t(do.call(data.frame, c(ag4, check.names = FALSE)))
d5 = data.frame(d4)
names(d5) = d4[1,]
d6 = d5[-1,]
xtable(d6,caption='Means and Standard Deviations for Relevant Variables by Regime.',digits=3, label='sumStats')
@
For each loan, Prosper offers 63 interesting variables, I now define an important subset.
\begin{itemize}
\item BorrowerRate: Annual interest rate paid by the borrower
\item TotalCreditLinespast7years: Number of credit lines, including cards and loans, used by the borrower in the seven years before the prosper loan was originated.
\item Inquirieslast6Months: Number of times a lender inquired about the Borrower's credit report, in the 6 months before the loan was originated. Traditional lenders make inquiries when a borrower applies for a loan, so this serves as a proxy for the number of times someone has borrowed in the past 6 months.
\item CurrentDelinquencies: Number of loans on which the borrower is delinquent at the origination date.
\item DelinquenciesLast7Years: The number of loans which the borrower has not repaid fully over the 7 years before the origination date.
\item MonthlyLoanPayment: The borrower's scheduled monthly payment on the Prosper loan.
\item CreditScoreRangeLower: The lower bound of the credit score range (which
  bounds his raw credit score) , that bounds a borrower's credit grade, as
  provided by Experian. Prosper provides the range instead of the raw score to
  preserve borrower anonymity.
\item CreditScoreEst: The average of the lower and upper bounds of the credit score range. 
\item PublicRecordsLast10Years: Number of public records of borrowing activity in the past 10 years at the origination date. 
\item DebtToIncomeRatio: The debt to income ratio of the borrower at the origination date. This value is capped at 10.01 (any debt to income ratio larger than 1000\% will be coded as 1001\%).
\item TotalProsperPaymentsBilled: Number of on time payments the borrower made on Prosper loans at the time they created the listing. This value is 0 if the borrower had no prior loans.
\item LenderYield: The lender's Yield on the loan. Equal to the Interest rate minus servicing fee.
\item BorrowerRate: The borrower's interest rate for the loan.
\item Investors: The number of investors that funded the loan.
\item *.sc: The Variable * was scaled to have 0 mean and unit variance. (found only in regression results)
\end{itemize}

@@

I do not discount or annualize ROIs, because in the period of low interest rates, the difference between default and repayment is large enough that a few percentage points don't matter: loans that default average -46\% ROI, loans that repay average + 24\% ROI.\footnote{Elaborate?}

\subsection{Prosper's Charged-Off vs. Default Terminology}
\footnote{remove this part?}
After the shutdown, Prosper began using the term ``Charged-Off" to mean that a loan is delinquent past 120 days. We observe more defaults than charge-offs before the shutdown, but more charge-offs than defaults after, likely because of the shift in terminology. According to Prosper's website:
\begin{quote}
In general, a debt or account is considered “charged off” when it is unlikely that further payments will be received. Debts are usually charged off after they remain unpaid for a period of time (e.g., 90 to 180 days). Prosper uses the 120 days as the charge off threshold because loans that become over 120 days past due are eligible for sale to a debt buyer, and we have found that there is a steep drop-off in likelihood of further payments after 120 days of delinquency.
\footnote{\url{http://blog.prosper.com/2008/09/16/charge-offs-explained/}}
\end{quote}

The rest of this paper documents my results, which are split into three sections.
Section 4 presents facts about the performance of loans across regimes which
suggest that early auction regime investors were not sufficiently sensitive to
risk, and that this partly explains the poor returns of the period. Section 5 tests and rejects
the EMH prediction that it should be impossible to earn excess returns
using publicly available information. Section 6 attempts to isolate the impact
of Prosper's policy changes by controlling for time period effects, and finds
that they explain at least some of the improvement in investor returns.

\section{Performance Across Regimes}
This section begins with an examination of ROIs over time and across
borrower credit buckets. I then discuss regime differences in default rates and
interest rates, the two features that largely determine returns.
The discussion of interest rates includes measurement of investors' varying sensitivity to risk
across regimes. The section closes by examining the impact of Prosper's interventions, which reduce the valid borrower pool to a less-risky subset and assign borrowers more informative credit grades, on investor returns.

@@

These findings suggests that although investors were extraordinarily unlucky in
the early auction regime, they made avoidably bad decisions that hurt their
returns. More specifically,
Finding (1) is inconsistent with the EMH prediction that risky assets like Prosper loans should offer investors a positive expected return.\footnote{Otherwise, investors would just hold cash.}
Finding (2) is inconsistent with the prediction that riskier loans should have higher expected returns to compensate investors for taking risks.
Finding (3) describes the sub-optimal investor decisions that led to (1) and (2), in response to arguments that investors chose an unlucky time period.
Finding (4) is inconsistent with Hayek's notion that a free market will arrive at accurate prices without the meddling of a central planner.

As the default data demonstrate, Finding (3) is only partially responsible for
Findings (1) and (2); investors were incredibly unlucky to have chosen to invest
in risky assets during turbulent macroeconomic times that many of the most
sophisticated investors did not see coming. In section 6, I attempt to control
for this time period effect to isolate the impact of the policy changes.
